---
title: "Coomba_trees_EDA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox/Personal/Coomba/CoombaTrees")
```

#Coomba trees
This document presents measurements of trees planted at Coomba by Mare and Mart Vesk.

##Read in the data

```{r read}
library(readxl)
d <- read_excel("~/Dropbox/Personal/Coomba/CoombaTrees/Data/Copy of coomba_trees_statistics_08012016.xlsx")
View(d)
summary(d)
```

```{r}
d$notes <- as.character(d$notes)
d$oh.can <- as.numeric((d$oh.can))
tail(d)
```

Remove some data, change some spp names. 'flg' to 'fg', 'bg' to 'sbg'
```{r}
d <- d[d$species!="exist frg",]
(ROWS <- d[d$species == 'bg', ] )
COLS <- 9
d[ROWS, COLS]
```


## Simple Plots

First, lets try to plot size through time:

```{r plot1, echo=FALSE, message=FALSE, warning=FALSE}
plot(d$girth_cm~d$years,  ylab = "Girth (cm)", xlab = "Years", pch=d$forest, cex=0.6, ylim = c(0,180), xlim = c(0,30))
```

Next, lets try to plot height through time:

```{r plot1, echo=FALSE, message=FALSE, warning=FALSE}
plot(d$ht_m~d$years, ylab = "Height (m)", pch=d$forest, cex=0.6, xlab = "Years", ylim = c(0,30), xlim = c(0,30))
```


and ht againast girth:
```{r allometry_plot, message=FALSE, warning=FALSE}
plot(d$ht_m~d$girth_cm,  ylab = "Height (m)", xlab = "Girth (cm)", pch=d$forest, cex=0.6, ylim = c(0,30), xlim = c(0,180)) 
legend(x = "topleft", c("new", "front", "east", "hill", "camp"), pch=c("n", "f", "E", "h", "C" ))
```
```{r subset by forest}
f<- d[d$forest=="hill_forest",]
plot(f$ht_m ~ f$girth_cm,  ylab = "Height (m)", xlab = "Girth (cm)", ylim = c(0,30), xlim = c(0,180))
```


## Recasting
get the girth data into wide format
```{r recast}
library(dplyr)
library(tidyr)
d_girth <-spread(d[,c(1,4,8,9,10)], years, girth_cm)
d_ht <-spread(d[,c(1,5,8,9,10)], years, ht_m)
```

Let's try plotting the trajectory of girth of individual trees.
```{r plot2, message=FALSE, warning=FALSE}
yrs <- c(min(d$years):23, 25:max(d$years))
matplot(x = yrs[], y = t(d_girth[,4:26]), type="o", cex=0.6,  ylab = "Girth (cm)", xlab = "Years",ylim = c(0,180), xlim = c(0,30))
```

Next, let's try the trajectory of height of individual trees.
```{r plot_ht_traj, message=FALSE, warning=FALSE}
matplot(x = yrs[], y = t(d_ht[,4:26]), type="o", cex=0.6, ylab = "Height (m)", xlab = "Years", ylim = c(0,30), xlim = c(0,30))
```

Lets go back and look at the competitive index 'oh.can'.

```{r}
hist(d$oh.can)
table(d$oh.can)
```

```{r}
plot(d$girth_cm ~ d$oh.can)
oh <- qlogis(0.8*(d$oh.can+0.125))
hist(oh)
```


# try some linear modeling with brms
```{r}
library(brms)
fit <- brm(count ~ log_Age_c + log_Base4_c * Trt_c + (1|patient) + (1|obs), 
           data = epilepsy, family = "poisson")
summary(fit, waic = TRUE) 
```

```{r}
plot(fit) 
```
Now nonlinear model, first on synthtic data:
```{r}
b <- c(2, 0.75)
x <- rnorm(100)
y <- rnorm(100, mean = b[1] * exp(b[2] * x))
dat1 <- data.frame(x, y)
```


```{r}
prior1 <- c(prior(normal(1, 2), nlpar = "b1"),
            prior(normal(0, 2), nlpar = "b2"))
fit1 <- brm(bf(y ~ b1 * exp(b2 * x), b1 + b2 ~ 1, nl = TRUE),
            data = dat1, prior = prior1)
```


Now with our trees. first a linear model

```{r}
fit <- brm(ht_m ~ years, #Hmax + scale * years, #+ (1|tree),
             #Hmax * (1 - exp(-log(2)/scale * year)) + (1|tree) + (1|forest),
          # Hmax ~ 1, scale ~ 1,
           data = d, family = gaussian())
      #     prior = c( prior(normal(40, 10), par = "Hmax"),
       #                   prior(normal(10, 5), par = "scale")) )

```


```{r}
summary(fit, waic = TRUE)
```


plot
```{r}
plot(marginal_effects(fit), points = TRUE)
```

Now a loglinear model:
```{r}
fit2 <- brm(ht_m ~ log(years), #Hmax + scale * years, #+ (1|tree),
             #Hmax * (1 - exp(-log(2)/scale * year)) + (1|tree) + (1|forest),
          # Hmax ~ 1, scale ~ 1,
           data = d, family = gaussian())
      #     prior = c( prior(normal(40, 10), par = "Hmax"),
       #                   prior(normal(10, 5), par = "scale")) )

```

```{r}
summary(fit2)
```

```{r}
plot(marginal_effects(fit2), points = TRUE)
```

Now a loglinear model with lognormal heights:
```{r}
fit3 <- brm(ht_m ~ log(years), #Hmax + scale * years, #+ (1|tree),
           data = d, family = lognormal() )
      #     prior = c( prior(normal(40, 10), par = "Hmax"),
       #                   prior(normal(10, 5), par = "scale")) )

```


```{r}
summary(fit3)
```

plot3
```{r}
plot(marginal_effects(fit3), points = TRUE)
```

Lets do some Posterior predictive checks

```{r}
pp_check(fit)
```

```{r}
pp_check(fit2)
```



```{r}
pp_check(fit3)
```
this overpredicts.drop the (semi) loglinear model with lognormal error. i think it is not meaningful and it fits poorly.

```{r}
LOO(fit, fit2, fit3, fit4, fit5, fit6, fit7)
```

so the loglinear model with gaussian errors is best fitting of linear models. Model three, which error increases with size, is poor. 
but log normal has a negative intercept, which is not right.

Perhaps the exponetnial model is best.

```{r}
pp_check(fit4)
```

Now a loglinear model with varying intercept and slope:
```{r}
fit5 <- brm(ht_m ~ log(years) + (log(years)|tree) + (log(years)|forest),
           data = d, family = gaussian(),
           control = list(adapt_delta = 0.9))
```

```{r}
summary(fit5)
```


```{r}
pp_check(fit5)
```
nice!

Now a loglinear model with varying intercept at tree and forest level but and slope at forest level:
```{r}
prior6 = c( prior(normal(10, 5), nlpar = "b1"),
           prior(normal(1, 1), nlpar = "b2"))
fit6 <- brm(ht_m ~ log(years) + (1|tree) + (log(years)|forest),
           data = d, family = gaussian(),
           control = list(adapt_delta = 0.9), cores = 2)
```

```{r}
summary(fit6)
```


```{r}
pp_check(fit6)
```
Also nice, but perhaps not perfect on the left side hump. try allowing varyaing slope for individuals, not intercept.

```{r}
prior7 = c( prior(normal(10, 5), nlpar = "b1"),
           prior(normal(1, 1), nlpar = "b2"))
fit7 <- brm(ht_m ~ log(years) + (0 + log(years)|tree) + (log(years)|forest),
           data = d, family = gaussian(),
           control = list(adapt_delta = 0.92, max_treedepth = 12), cores = 2)
```


```{r}
pp_check(fit7)
```


```{r nonlinear height growth}
fit4 <- brm(bf(ht_m ~ Hmax * (1 - exp(-log(2) / scale * years) ) , 
Hmax ~ 1 +  (1|forest), scale ~ 1, nl = TRUE),

data = d, family = gaussian(),
prior = c(prior(student_t(3, 40, 10), lb=0, nlpar = "Hmax"),
prior(student_t(3, 20, 10), lb=0, nlpar = "scale")),
                control = list(adapt_delta = 0.9))

```


```{r}
plot(fit4)
```



```{r}
summary(fit4)
```


```{r}
pp_check(fit4)
```

plot
```{r}
plot(marginal_effects(fit4), points = TRUE)
```

facet plot
```{r}
conditions <- data.frame(forest = unique(d$forest))
rownames(conditions) <- unique(d$forest)
plot(marginal_effects(fit4, conditions = conditions, 
                      re_formula = NULL, method = "predict"), 
     ncol = 2, points = TRUE)
```


